---
title: "`valr`: Genome interval arithmetic in R"
author: Jay Hesselberth <jay.hesselberth@gmail.com>
output: github_document
---

[![Build Status](https://travis-ci.com/jayhesselberth/valr.svg?token=Q9WRSyqYnpS7KpFfTscp&branch=master)](https://travis-ci.com/jayhesselberth/valr)
[![Coverage Status](https://img.shields.io/codecov/c/github/jayhesselberth/valr/master.svg)](https://codecov.io/github/jayhesselberth/valr?branch=master)

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

`valr` provides methods to do interval manipulations **within the R environment**, enabling fast explorative analysis of genome-scale data.

# Installation

`valr` can be installed from github:

```r
> devtools::install_github('jayhesselberth/valr')
```

# Overview

`valr` enables the analysis of genome intervals in R. To keep the package lightweight, `valr` only supports BED+ and bedGraph formats. Key parts are implemented in `Rcpp` for speed.

The goal of `valr` is to enable easy analysis of genome-scale data sets **within R**. Moreover, `valr` makes use of new R libraries like `dplyr` and its pipe operator(`%>%`) for an expressive syntax that makes genome analysis fun. So a workflow like [this](https://github.com/arq5x/bedtools-protocols/blob/master/bedtools.md#bp3-plot-transcription-factor-occupancy-surrounding-the-transcription-start-site) becomes:

```{r tss_signal_example, warning = FALSE, message = FALSE}
library(valr)
library(dplyr)
library(ggplot2)

bedfile <- system.file('extdata', 'genes.hg19.chr22.bed.gz', package = 'valr')
bgfile  <- system.file('extdata', 'hela.h3k4.chip.bg.gz', package = 'valr')
genomefile <- system.file('extdata', 'hg19.chrom.sizes.gz', package = 'valr')

genes <- read_bed(bedfile, n_fields = 6)
genome <- read_genome(genomefile)

tss <- genes %>% filter(strand == '+') %>% mutate(end = start + 1)

x <- tss %>%
  bed_slop(genome, both = 1000) %>%
  bed_makewindows(genome, win_size = 50)

y <- read_bedgraph(bgfile)

res <- bed_map(x, y, add_group = 'win_id.x', sums = sum(value.y)) %>%
  group_by(win_id.x) %>%
  summarize(means = mean(sums), sds = sd(sums))

limits <- aes(ymax = means + sds, ymin = means - sds)
ggplot(res, aes(x = win_id.x, y = means)) +
  geom_point()  + geom_pointrange(limits)
```

# Vignette

See the vignettes for a full description of the package and examples (`browseVignettes(package = "valr")`)

