---
title: 'valr overview'
output:
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteIndexEntry{valr-overview}
      %\VignetteEncoding{UTF-8}
      %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r knitr_opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

```{r init, echo = FALSE, message = FALSE}
library(valr)
library(dplyr)
library(ggplot2)
library(tibble)
```

## valr Overview

Functions in `valr` have similar names to their `BEDtools` counterparts,
and so will be familiar to users coming from the `BEDtools` suite.

Similar to
[`pybedtools`](https://daler.github.io/pybedtools/#why-pybedtools),
`valr` has a terse syntax:

```{r syntax_demo, message = FALSE}
library(valr)
library(dplyr)

snps <- read_bed(valr_example('hg19.snps147.chr22.bed.gz'), n_fields = 6)
genes <- read_bed(valr_example('genes.hg19.chr22.bed.gz'), n_fields = 6)

# find snps in intergenic regions
intergenic <- bed_subtract(snps, genes)
# distance from intergenic snps to nearest gene
nearby <- bed_closest(intergenic, genes)

nearby %>%
  select(starts_with('name'), .overlap, .dist) %>%
  filter(abs(.dist) < 1000)
```

### Input data

`valr` assigns common column names to facilitate comparisons between
tbls. All tbls will have `chrom`, `start`, and `end` columns, and some
tbls from multi-column formats will have additional pre-determined
column names. See the `read_bed()` documentation for details.

```{r file_io}
bed_file <- valr_example("3fields.bed.gz")
read_bed(bed_file) # accepts filepaths or URLs
```

`valr` can also operate on BED-like data.frames already constructed in
R, provided that columns named `chrom`, `start` and `end` are present.
New tbls can also be constructed as either `tibbles` or base R
`data.frames`.

```{r trbl_ivls}
bed <- tribble(
  ~chrom, ~start,  ~end, 
  "chr1", 1657492, 2657492, 
  "chr2", 2501324, 3094650
)

bed
```

### Interval coordinates

`valr` adheres to [BED
format](https://genome.ucsc.edu/FAQ/FAQformat#format1), which specifies
that the start position for an interval is zero based and the end
position is one-based. The first position in a chromosome is 0. The end
position for a chromosome is one position passed the last base, and is
not included in the interval. For example:

```{r zero-based}
# a chromosome 100 basepairs in length
chrom <- tribble(
  ~chrom, ~start, ~end, 
  "chr1", 0,      100
)

chrom

# single base-pair intervals
bases <- tribble(
  ~chrom, ~start, ~end, 
  "chr1", 0,      1, # first base of chromosome
  "chr1", 1,      2,  # second base of chromosome
  "chr1", 99,     100 # last base of chromosome
)

bases
```

### Remote databases

Remote databases can be accessed with `db_ucsc()` (to access the UCSC
Browser) and `db_ensembl()` (to access Ensembl databases).

```{r db, warning = FALSE, eval = FALSE}
# access the `refGene` tbl on the `hg38` assembly.
if(require(RMariaDB)) {
  ucsc <- db_ucsc('hg38')
  tbl(ucsc, 'refGene')
}
```

### Visual documentation

The `bed_glyph()` tool illustrates the results of operations in `valr`,
similar to those found in the `BEDtools` documentation. This glyph shows
the result of intersecting `x` and `y` intervals with `bed_intersect()`:

```{r intersect_glyph}
x <- tribble(
  ~chrom, ~start, ~end,
  'chr1', 25,     50,
  'chr1', 100,    125
)

y <- tribble(
  ~chrom, ~start, ~end,
  'chr1', 30,     75
)

bed_glyph(bed_intersect(x, y))
```

And this glyph illustrates `bed_merge()`:

```{r merge_glyph}
x <- tribble(
  ~chrom, ~start, ~end,
  'chr1',      1,      50,
  'chr1',      10,     75,
  'chr1',      100,    120
)

bed_glyph(bed_merge(x))
```

### Grouping data

The `group_by` function in dplyr can be used to perform functions on
subsets of single and multiple `data_frame`s. Functions in `valr`
leverage grouping to enable a variety of comparisons. For example,
intervals can be grouped by `strand` to perform comparisons among
intervals on the same strand.

```{r strand}
x <- tribble(
  ~chrom, ~start, ~end, ~strand,
  'chr1', 1,      100,  '+',
  'chr1', 50,     150,  '+',
  'chr2', 100,    200,  '-'
)

y <- tribble(
  ~chrom, ~start, ~end, ~strand,
  'chr1', 50,     125,  '+',
  'chr1', 50,     150,  '-',
  'chr2', 50,     150,  '+'
)

# intersect tbls by strand
x <- group_by(x, strand)
y <- group_by(y, strand)

bed_intersect(x, y)
```

Comparisons between intervals on opposite strands are done using the
`flip_strands()` function:

```{r strand_opp}
x <- group_by(x, strand)

y <- flip_strands(y)
y <- group_by(y, strand)

bed_intersect(x, y)
```

Both single set (e.g. `bed_merge()`) and multi set operations will
respect groupings in the input intervals.

### Column specification

Columns in `BEDtools` are referred to by position:

``` bash
# calculate the mean of column 6 for intervals in `b` that overlap with `a`
bedtools map -a a.bed -b b.bed -c 6 -o mean
```

In `valr`, columns are referred to by name and can be used in multiple
name/value expressions for summaries.

```{r NSE, eval = FALSE}
# calculate the mean and variance for a `value` column
bed_map(a, b, .mean = mean(value), .var = var(value))

# report concatenated and max values for merged intervals
bed_merge(a, .concat = concat(value), .max = max(value))
```

## Getting started

### Meta-analysis

This demonstration illustrates how to use `valr` tools to perform a
"meta-analysis" of signals relative to genomic features. Here we to
analyze the distribution of histone marks surrounding transcription
start sites.

```{r demo-tss, warning = FALSE, message = FALSE}
# `valr_example()` identifies the path of example files
bedfile <- valr_example('genes.hg19.chr22.bed.gz')
genomefile <- valr_example('hg19.chrom.sizes.gz')
bgfile  <- valr_example('hela.h3k4.chip.bg.gz')

genes <- read_bed(bedfile, n_fields = 6)
genome <- read_genome(genomefile)
y <- read_bedgraph(bgfile)
```

Then we generate 1 bp intervals to represent transcription start sites
(TSSs). We focus on `+` strand genes, but `-` genes are easily
accommodated by filtering them and using `bed_makewindows()` with
`reversed` window numbers.

```{r tss}
# get transcription start site (TSS) intervals, `+` strand only
tss <- create_tss(genes) %>%
  filter(strand == '+')

# look 1000 bp up and downstream
region_size <- 1000
# in 50 bp windows
win_size <- 50

# add slop to the TSS, break into windows
x <- tss %>%
  bed_slop(genome, both = region_size) %>%
  bed_makewindows(win_size)

# now have `.win_id` for grouped calculations
x
```

Now we use the `.win_id` group with `bed_map()` to calculate a sum by
mapping `y` signals onto the intervals in `x`. These data are regrouped
by `.win_id` and a summary with `mean` and `sd` values is calculated.

```{r map}
# map signals to TSS regions and calculate summary statistics for each `.win_id`.
res <- bed_map(
  x, y,
  win_sum = sum(value, na.rm = TRUE)
  ) %>%
  group_by(.win_id) %>%
  summarize(
    win_mean = mean(win_sum, na.rm = TRUE),
    win_sd = sd(win_sum, na.rm = TRUE)
  )

res
```

Finally, these summary statistics are used to construct a plot that
illustrates histone density surrounding TSSs.

```{r plot, warning = FALSE, message = FALSE, fig.align='center', fig.width=6}
library(ggplot2)

x_labels <- seq(-region_size, region_size, by = win_size * 5)
x_breaks <- seq(1, 41, by = 5)

sd_limits <- aes(ymax = win_mean + win_sd, ymin = win_mean - win_sd)

ggplot(
  res,
  aes(x = .win_id, y = win_mean)
) +
  geom_point() +
  geom_line() +
  geom_pointrange(sd_limits, alpha = 0.4) + 
  scale_x_continuous(
    labels = x_labels, breaks = x_breaks
    ) + 
  labs(
    x = 'Position (bp from TSS)',
    y = 'Signal',
    title = 'Human H3K4me3 signal near transcription start sites'
  ) +
  theme_classic()
```

## Related work

Several tools implement functionality similar to valr with different
interfaces.

#### Command-line tools

-   [BEDtools](https://bedtools.readthedocs.io/en/latest/)
-   [bedops](https://bedops.readthedocs.io/en/latest/index.html).

#### Python

-   [pybedtools](https://pythonhosted.org/pybedtools/)

#### R packages

-   [GenomicRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html)
-   [bedr](https://CRAN.R-project.org/package=bedr)
-   [IRanges](https://bioconductor.org/packages/release/bioc/html/IRanges.html)
-   [GenometriCorr](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002529)
