---
title: 'valr overview'
author: "Jay Hesselberth <jay.hesselberth@gmail.com>"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    vignette: >
      %\VignetteIndexEntry{valr}
      %\VignetteEngine{knitr::rmarkdown}
      %\VignetteEncoding{UTF-8}
---

```{r knitr_opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/overview-",
  fig.height = 3,
  fig.align = "center",
  fig.width = 4
)
```

```{r init, echo = FALSE, message = FALSE}
library(valr)
library(dplyr)
library(ggplot2)
library(tibble)
```
## Why `valr`?

**Why another tool set for interval manipulations?** There are several [other software packages](#related_work) available for genome interval analysis. However, based on our experiences teaching genome analysis, we were motivated to develop a toolset that:

* Combines analysis and visualization in RStudio.
* Can be used to generate reports with Rmarkdown.
* Is highly extensible. New tools are quickly implemented on the R side.
* Leverages the "modern R" syntax, using `dplyr` and the pipe operator from `magrittr` (`%>%`).
* Maximizes speed by implementing compute-intensive algorithms in `Rcpp`.
* Facilitates interactive visulaizations with [`shiny`][13].

`valr` can currently be used for analysis of pre-processed data in BED and related formats. We plan to support BAM and VCF files soon via tabix indexes.

### Familiar tools, all within R

The functions in `valr` have similar names to their `BEDtools` counterparts, and so will be familiar to users coming from the `BEDtools` suite. Similar to [`pybedtools`](https://daler.github.io/pybedtools/#why-pybedtools), `valr` has a terse syntax:

```{r syntax_demo, message = FALSE}
library(valr)
library(dplyr)

snps <- read_bed(valr_example('hg19.snps147.chr22.bed.gz'), n_fields = 6)
genes <- read_bed(valr_example('genes.hg19.chr22.bed.gz'), n_fields = 6)

# find snps in intergenic regions
intergenic <- bed_subtract(snps, genes)
# distance from intergenic snps to nearest gene
nearby <- bed_closest(intergenic, genes)

nearby %>%
  select(starts_with('name'), .overlap, .dist) %>%
  filter(abs(.dist) < 1000)
```

### Remote databases

Remote databases can be accessed with `db_ucsc()` (to access the UCSC Browser) and `db_ensembl()` (to access Ensembl databases).

```{r db}
# access the `refGene` tbl on the `hg38` assembly.
ucsc <- db_ucsc('hg38')
tbl(ucsc, 'refGene')
```

### Visual documentation

The `bed_glyph()` tool illustrates the results of operations in `valr`, similar to those found in the `BEDtools` documentation. This glyph shows the result of intersecting `x` and `y` intervals with `bed_intersect()`:

```{r intersect_glyph}
x <- tibble::tribble(
  ~chrom, ~start, ~end,
  'chr1', 25,     50,
  'chr1', 100,    125
)

y <- tibble::tribble(
  ~chrom, ~start, ~end,
  'chr1', 30,     75
)

bed_glyph(bed_intersect(x, y))
```

And this glyph illustrates `bed_merge()`:

```{r merge_glyph}
x <- tibble::tribble(
  ~chrom, ~start, ~end,
  'chr1',      1,      50,
  'chr1',      10,     75,
  'chr1',      100,    120
)

bed_glyph(bed_merge(x))
```

### Reproducible reports 

`valr` can be used in RMarkdown documents to generate reproducible work-flows for data processing. Because `valr` is reasonably fast (see the [benchmarks](#benchmarks)), we now use it in lieu of other tools for exploratory analysis of genomic data sets in R.

Command-line tools like `BEDtools` and `bedops` can be used in reproducible workflows (e.g., with [`snakemake`][12]), but it is cumbersome to move from command-line tools to exploratory analysis and plotting software. [`pybedtools`][4] can be used within `ipython notebooks` to accomplish a similar goal, but others have pointed out [issues with this approach][11], including clunky version control. Because RMarkdown files are text files, they are readily kept under version control. Moreover, new features in RStudio (e.g. notebook viewing) enable similar functionality to `ipython`.

### Grouping data

The `group_by` function in dplyr can be used to perform fuctions on subsets of single and multiple `data_frame`s. Functions in `valr` leverage grouping to enable a variety of comparisons. For example, intervals can be grouped by `strand` to perform comparisons among intervals on the same strand.

```{r strand}
x <- tibble::tribble(
  ~chrom, ~start, ~end, ~strand,
  'chr1', 1,      100,  '+',
  'chr1', 50,     150,  '+',
  'chr2', 100,    200,  '-'
)

y <- tibble::tribble(
  ~chrom, ~start, ~end, ~strand,
  'chr1', 50,     125,  '+',
  'chr1', 50,     150,  '-',
  'chr2', 50,     150,  '+'
)

# intersect tbls by strand
x <- group_by(x, strand)
y <- group_by(y, strand)

bed_intersect(x, y)
```

Comparisons between intervals on opposite strands are done using the `flip_strands()` function:

```{r strand_opp}
x <- group_by(x, strand)

y <- flip_strands(y)
y <- group_by(y, strand)

bed_intersect(x, y)
```

Both single set (e.g. `bed_merge()`) and multi set operations will respect groupings in the input intervals.

### Column specification

Columns in `BEDtools` are referred to by position:

```bash
# calculate the mean of column 6 for intervals in `b` that overlap with `a`
bedtools map -a a.bed -b b.bed -c 6 -o mean
```

In `valr`, columns are referred to by name and can be used in multiple name/value expressions for summaries.

```{r NSE, eval = FALSE}
# calculate the mean and variance for a `value` column
bed_map(a, b, .mean = mean(value), .var = var(value))

# report concatenated and max values for merged intervals
bed_merge(a, .concat = concat(value), .max = max(value))
```

## Getting started

### Meta-analysis

This demonstration illustrates how to use `valr` tools to perform a "meta-analysis" of signals relative to genomic features. Here we to analyze the distribution of histone marks surrounding transcription start sites.

First we load libraries and relevant data.

```{r demo-tss, warning = FALSE, message = FALSE}
# `valr_example()` identifies the path of example files
bedfile <- valr_example('genes.hg19.chr22.bed.gz')
genomefile <- valr_example('hg19.chrom.sizes.gz')
bgfile  <- valr_example('hela.h3k4.chip.bg.gz')

genes <- read_bed(bedfile, n_fields = 6)
genome <- read_genome(genomefile)
y <- read_bedgraph(bgfile)
```

Then we generate 1 bp intervals to represent transcription start sites (TSSs). We focus on `+` strand genes, but `-` genes are easily accomodated by filtering them and using `bed_makewindows()` with `reverse`d window numbers.

```{r tss}
# generate 1 bp TSS intervals, `+` strand only
tss <- genes %>%
  filter(strand == '+') %>%
  mutate(end = start + 1)

# 1000 bp up and downstream
region_size <- 1000
# 50 bp windows
win_size <- 50

# add slop to the TSS, break into windows and add a group
x <- tss %>%
  bed_slop(genome, both = region_size) %>%
  bed_makewindows(genome, win_size)

x
```

Now we use the `.win_id` group with `bed_map()` to caluclate a sum by mapping `y` signals onto the intervals in `x`. These data are regrouped by `.win_id` and a summary with `mean` and `sd` values is calculated.

```{r map}
# map signals to TSS regions and calculate summary statistics.
res <- bed_map(x, y, win_sum = sum(value, na.rm = TRUE)) %>%
  group_by(.win_id) %>%
  summarize(win_mean = mean(win_sum, na.rm = TRUE),
            win_sd = sd(win_sum, na.rm = TRUE))

res
```

Finally, these summary statistics are used to construct a plot that illustrates histone density surrounding TSSs.

```{r plot, warning = FALSE, message = FALSE, fig.align='center', fig.width=6}
library(ggplot2)

x_labels <- seq(-region_size, region_size, by = win_size * 5)
x_breaks <- seq(1, 41, by = 5)

sd_limits <- aes(ymax = win_mean + win_sd, ymin = win_mean - win_sd)

ggplot(res, aes(x = .win_id, y = win_mean)) +
  geom_point() + geom_pointrange(sd_limits) + 
  scale_x_continuous(labels = x_labels, breaks = x_breaks) + 
  xlab('Position (bp from TSS)') + ylab('Signal') + 
  ggtitle('Human H3K4me3 signal near transcription start sites') +
  theme_classic()
```

### Interval statistics

Estimates of significance for interval overlaps can be obtained by combining `bed_shuffle()`, `bed_random()` and the `sample_` functions from `dplyr` with interval statistics in `valr`.

Here we examine the overlap of repeat classes in the human genome (on `chr22` only, for simplicity) using `bed_jaccard()`.

```{r reldist_shuffle, message = FALSE, warning = FALSE}
library(purrr)
library(tidyr)

repeats <- read_bed(valr_example('hg19.rmsk.chr22.bed.gz'), n_fields = 6) 
genome <- read_genome(valr_example('hg19.chrom.sizes.gz'))

shuffle_intervals <- function(n, .data, genome) {
  replicate(n, bed_shuffle(.data, genome), simplify = FALSE) %>%
    bind_rows(.id = 'rep') %>%
    group_by(rep) %>% nest()
}

shuffled <- shuffle_intervals(n = 100, repeats, genome) %>%
  mutate(jaccard = data %>%
           map(bed_jaccard, repeats) %>%
           map_dbl("jaccard"))
  
shuffled
```

## Benchmarks

Certain algorithms in `valr` were implemented in `Rcpp` to enable fluid interactive analysis.  This graph illustrates the timing of functions in `valr` for two sets of 100,000 random 1 kilobase intervals from the hg19 genome.

```{r benchmarks, echo = FALSE, fig.width=6}
library(microbenchmark)

genome <- read_genome(valr_example('hg19.chrom.sizes.gz'))

x <- bed_random(genome, n=1e5)
y <- bed_random(genome, n=1e5)

ts <- microbenchmark(
  bed_random(genome), bed_closest(x, y),
  bed_intersect(x, y), bed_merge(x),
  bed_subtract(x, y), bed_complement(x, genome),
  bed_shuffle(x, genome),
  times = 1,
  unit = 's')

# from unexported microbenchmark::convert_to_unit
ts$ntime <- ts$time / 1e9 

ggplot(ts, aes(y=reorder(expr, ntime), x=ntime)) +
  geom_point(color='red', size=4) +
  xlab('execution time (seconds)') + ylab('') +
  theme_bw()
```
 
## API

Function names are similar to their their [BEDtools][1] counterparts, with some additions.

### Reading data

* BED and related files are read with `read_bed()`, `read_bed12()`, `read_bedgraph()`, `read_narrowpeak()` and `read_broadpeak()`.
  
* Genome files containing chromosome name and size information are loaded with `read_genome()`.
  
* VCF files are loaded with `read_vcf()`.

### Transforming single interval sets

* Intervals are ordered with `bed_sort()`.

* Interval coordinates are adjusted with `bed_slop()` and `bed_shift()`, and new flanking intervals are created with `bed_flank()`.

* Nearby intervals are combined with `bed_merge()` and identified (but not merged) with `bed_cluster()`.  

* Intervals not covered by a query are created with `bed_complement()`.

### Comparing multiple interval sets

* Find overlaps between two sets of intervals with `bed_intersect()`.

* Apply functions to selected columns for overlapping intervals with `bed_map()`.

* Remove intervals based on overlaps between two files with `bed_subtract()`.

* Find overlapping intervals within a window with `bed_window()`.

* Find the closest intervals independent of overlaps with `bed_closest()`.

### Randomizing intervals

* Generate random intervals from an input genome with `bed_random()`.

* Shuffle the coordinates of input intervals with `bed_shuffle()`.

* Random sampling of input intervals is done with the `sample_` function family in `dplyr`.

### Interval statistics

* Quantify overlaps between two sets of intervals with `bed_fisher()`.

* Quantify relative and absolute distances between sets of intervals with `bed_reldist()` and `bed_absdist()`.

* Quantify extent of overlap between two sets of intervals with `bed_jaccard()` or `bed_projection()`.

### Utilities

* Visualize the actions of valr functions with `bed_glyph()`.

* Constrain intervals to a genome reference with `bound_intervals()`.

* Subdivide intervals with `bed_makewindows()`.

* Convert BED12 to BED6 format with `bed12_to_exons()`.

* Calculate spacing between intervals with `interval_spacing()`.

## Related work {#related_work}

* Command-line tools [BEDtools][1] and [bedops][5].

* The Python library [pybedtools][4] wraps BEDtools.

* The R packages [GenomicRanges][6], [bedr][7], [IRanges][8] and [GenometriCorr][9] provide similar capability with a different philosophy.

[1]: http://bedtools.readthedocs.org/en/latest/
[2]: https://github.com/hadley/dplyr
[3]: http://www.rcpp.org/
[4]: https://pythonhosted.org/pybedtools/
[5]: http://bedops.readthedocs.org/en/latest/index.html
[6]: https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html
[7]: https://CRAN.R-project.org/package=bedr
[8]: https://bioconductor.org/packages/release/bioc/html/IRanges.html
[9]: http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002529
[10]: http://rmarkdown.rstudio.com/
[11]: https://www.r-bloggers.com/why-i-dont-like-jupyter-fka-ipython-notebook/
[12]: https://bitbucket.org/snakemake/snakemake/wiki/Home
[13]: http://shiny.rstudio.com/
